A Monte Carlo Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-07-18 doc/jcomp2012/jcomp2012.tex
Almost done with rough draft for jcomp paper

\newcommand{\vA}{\ensuremath{\ve{A}}}
\newcommand{\vb}{\ensuremath{\ve{b}}}
\newcommand{\vx}{\ensuremath{\ve{x}}}
\newcommand{\vr}{\ensuremath{\ve{r}}}
\newcommand{\vI}{\ensuremath{\ve{I}}}
\newcommand{\vH}{\ensuremath{\ve{H}}}
\newcommand{\vP}{\ensuremath{\ve{P}}}

@@

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equations.  We show
    that this method can be an effective solver for sparse matrix
    systems.  We also discuss its general applicability to broader
    classes of problems.

@@

  \begin{keyword}
    radiation diffusion \sep Monte Carlo \sep sparse matrix systems
  \end{keyword}

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent discrete systems.  In this
paper, we extend our previous work in the following manner:
\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this solver to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}
In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
Monte Carlo to solve the discrete radiation diffusion equation in
\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
given in \S~\ref{sec:results}.  We finish with conclusions and ideas
for future work in \S~\ref{sec:conclusions}.

@@

Consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\
    &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
spectral radius, $\rho(\vH)$, of the iteration matrix satisfies
\begin{equation}
  \rho(\vH) < 1\:,
\end{equation}
we can expand $\vA^{-1}$ using the \latin{Neumann Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\
    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.

@@

Now we consider a Monte Carlo method that can be used to estimate the solution
to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First, rewrite
Eq.~(\ref{eq:neumann_series}) in order to calculate a component of
$\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
This series can be interpreted as a series of transitions from
$i_{m-1}\rightarrow i_m$ that can be simulated by a random walk.
Let $X$ be a random variable sampled from a random walk with $k$
events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. When the particle
transitions between states, for example $i\rightarrow j$, the weight
is multiplied by the factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:weight}
\end{equation}
Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
    &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.
Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
estimator of the components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities, $p_{ij}$. The most
straightforward approach is to set
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}
Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function for selecting a new state $j$,
given that the current state is $i$.  Random walks can be terminated
in two ways: the matrix can be augmented with a terminating event
equation that describes the probability that a particle ends its walk,
or the random walk can be terminated by weight cutoff.  Generally, we
choose to terminate random walks using a weight cutoff as opposed to
augmenting the matrix.

@@

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk.  In
the mathematical literature this is referred to as an adjoint method;
however, to transport practitioners this approach more closely
resembles a traditional forward method.  In this method, the weight
change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source in cell/state $i_0$.  The
Kronecker delta implies that tallies are only made in the cell/state
where the random walk currently resides.  This is the common approach
in standard Monte Carlo transport simulations.  We will investigate
this approach as applied to the 1T thermal radiation diffusion
problem in \S~\ref{sec:solut-radi-diff}.

Similar to the Direct Method, the Adjoint Method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk, $W_c$ is the
input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
weight of the source particle.  The random walk ends on the
$m^\text{th}$ step if $W_m < W_f$.

@@

The methods presented in \S~\ref{sec:monte-carlo-matrix} are
characterized by slow convergence rates.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual
scheme called Sequential Monte Carlo to speed up the convergence of
these methods.  A variant of this scheme has been successfully applied
to the 1D nonlinear thermal radiation diffusion equation in
Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:
\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is
iterated to convergence.  This iteration sequence is closely related
to iterative refinement; the exception being that there is no update
to to $x^{l+1}$ between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration scheme for fixed-point iteration. The
Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method is
defined as follows:
\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:, \\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,\\
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %% 
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:general-MCSA}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method.
The error is used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence
based on the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

\subsection{Monte Carlo Method Selection and Convergence}

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method. To demonstrate the effectiveness
of the adjoint method over the direct method, we choose the
two-dimensional time-dependent Poisson equation as a simple model
problem

\begin{equation}
  \frac{\partial u}{\partial t} = \nabla^2 u
  \label{eq:poisson_equation}
\end{equation}

@@

\begin{equation}
  \nabla^2_5 = \frac{1}{h^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil both assuming a grid size of $h$
in both the $i$ and $j$ directions.

\begin{equation}
  \nabla^2_9 = \frac{1}{6h^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4 u_{i,j-1}
    + 4 u_{i,j+1} + u_{i-1,j-1} + u_{i-1,j+1} + u_{i+1,j-1} +
    u_{i+1,j+1} - 20 u_{i,j}]
  \label{eq:nine_point_stencil}
\end{equation}

For a single time step solution, we then have the following linear
system to be solved with the MCSA method.

\begin{equation}
  A u^{n+1} = u^n
  \label{eq:poisson_eq_lin_sys}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant grid size and was
used with varying values of problem size, fixing the spectral radius
of the system at a constant value for each variation. Both the five
point and nine point stencils were used with both the direct and
adjoint solvers. For each case, 50 histories were simulated per cycle
with a weight cutoff of $1.0E-4$ for the stochastic solver and a
convergence tolerance of $1.0E-8$ for the MCSA solver. All
computations were completed on a 2.66 GHz Intel Core i7 machine with 8
GB 1067 MHz DDR3 memory.

Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size. We see clearly in
Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver with
MCSA results in several orders of magnitude speedup over the direct
solver while the number of iterations required to converge is of a
similar scale. We expect this for several reasons. First, with an
equivalent number of histories specified for both solvers and a system
of size $N$, the direct solver will compute $N$ times the number of
histories random walks while the adjoint solver will only compute the
number of histories specified. Initially, this may not seem like a
fair comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of histories per cycle is in fact larger for the direct method,
the solution computed at each iteration is no better than that
computed by the adjoint method.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    mesh). Both the adjoint and direct solvers are used with the five
    point and nine point stencils. A significant CPU time speedup is
    noted with the adjoint method due to the reduced number of
    histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
    mesh. Both the adjoint and direct solvers are used with the five
    point and nine point stencils. Both methods give similar iteration
    behavior.}
  \label{fig:poisson_iterations}
\end{figure}

Although CPU time is shown to be greatly decreased by using the
adjoint solver in Fig.~\ref{fig:poisson_cpu_time}, the iteration
behavior was approximately the same. To assess the convergence
properties of MCSA using each solver and stencil, the infinity norm of
the residual computed in Eq.~(\ref{eq:general-MCSA}) was computed at
each iteration for a fixed problem
size. Fig.~\ref{fig:poisson_convergence} gives the results of these
computations. First, it is worthy to note on the semilog plot that we
are indeed achieving the expected exponential convergence from
MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior.

@@

We have presented a new Monte Carlo solution method for solving the
discrete, time-dependent diffusion equation in 3D.  The MCSA method
has been shown to match results using standard solution techniques to
arbitrary precision.  Also, the new method is faster than
preconditioned CG and Richardson iteration.

>>>>

\newcommand{\vA}{\ensuremath{\ve{A}}}
\newcommand{\vb}{\ensuremath{\ve{b}}}
\newcommand{\vx}{\ensuremath{\ve{x}}}
\newcommand{\vr}{\ensuremath{\ve{r}}}
\newcommand{\vu}{\ensuremath{\ve{u}}}
\newcommand{\vI}{\ensuremath{\ve{I}}}
\newcommand{\vH}{\ensuremath{\ve{H}}}
\newcommand{\vP}{\ensuremath{\ve{P}}}

@@

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation.  We show that
    this method can be an effective solver for sparse matrix systems
    and is competitive with deterministic methods.  We also discuss
    its general applicability to broader classes of problems.

@@

  \begin{keyword}
    radiation diffusion \sep synthetic acceleration \sep Monte Carlo
    \sep sparse matrix systems
  \end{keyword}

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent discrete systems. In this
work, we extend our previous work in the following manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this solver to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
Monte Carlo to solve the discrete radiation diffusion equation in
\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
given in \S~\ref{sec:results}.  We finish with conclusions and ideas
for future work in \S~\ref{sec:conclusions}.

@@

Consider the following matrix equation:

\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}

which can be written,

\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\
    &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}

Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
spectral radius, $\rho(\vH)$, of the iteration matrix satisfies

\begin{equation}
  \rho(\vH) < 1\:,
\end{equation}

we can expand $\vA^{-1}$ using the \latin{Neumann Series},

\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}

Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,

\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\
    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}

With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),

\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}

Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,

\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}

The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events (particles) that initiates in state $i$,

\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}

Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. At every step,
contributions to each component of $X$ are provided only by the
starting state of the particle, $b_{i_m}$. Therefore, for a given
random walk permutation, each state must be used as a starting value
in order to calculated its contribution to the expectation value
(i.e. a state of size $N$ will have $N$ particles per random walk
permutation). When the particle transitions between states, for
example $i\rightarrow j$, the particle weight is multiplied by the
transition factor.

\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}

where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is

\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
    &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}

where $\nu$ denotes a particular random walk permutation.
Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
estimator of the components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities, $p_{ij}$. The most
straightforward approach is to set

\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}

Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function that can be sampled to select a
new state $j$, given that the current state is $i$.  Random walks can
be terminated in two ways: the matrix can be augmented with a
terminating event equation that describes the probability that a
particle ends its walk, or the random walk can be terminated by weight
cutoff, $W_c$.  Generally, we choose to terminate random walks using a
weight cutoff as opposed to augmenting the matrix sucth that the
random walk terminates when $W_m < W_c$.

@@

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk.  In
the mathematical literature this is referred to as an adjoint method;
however, to transport practitioners this approach more closely
resembles a traditional forward method.  In this method, the weight
change from state $i\rightarrow j$ is

\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}

The transition probabilities may be calculated as

\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}

Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is

\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}

Here, $\hat{b}_{i_0}$ is the sampled source and initial particle
weight in state $i_0$.  The Kronecker delta implies that tallies are
only made in the state where the random walk currently resides.  This
is the common approach in standard Monte Carlo transport
simulations. Unlike the direct method, because tallies are made in the
state in which the particle currently resides, we are no longer
required to start a particle in each state for each random walk
permutation. Instead, sampling the source is suffucient and requires
only one particle per permutation.

Similar to the Direct Method, the Adjoint Method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that

\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}

where $W_f$ is the terminating weight of the random walk, $W_c$ is the
input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
weight of the source particle.  The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

@@

The methods presented in \S~\ref{sec:monte-carlo-matrix} are
characterized by slow convergence rates bound by the Central Limit
Theorem.  Halton~\cite{halton_1962,halton_1994} proposed a staged
residual scheme called Sequential Monte Carlo to speed up the
convergence of these methods.  A variant of this scheme has been
successfully applied to the 1D nonlinear thermal radiation diffusion
equation in Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:

\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}

In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is
iterated to convergence.  This iteration sequence is closely related
to iterative refinement; the exception being that there is no update
to to $x^{l+1}$ between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration scheme for fixed-point iteration. We begin by
subtracting Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):

\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}

where

\begin{equation}
  \delta \vx^{l+1} = \vx - \vx^{l+1}
  \label{eq:lplus1error}
\end{equation}

is defined as the error at iteration $l+1$. We then subtract
$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}) giving:

\begin{equation}
  \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
  \label{eq:synthetic_error2}
\end{equation}

The following iterative scheme will then converge in one iteration:

\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
    \label{eq:synthetic-residual_solve}\\ 
    %% 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
  \label{eq:general-synthetic}
\end{subequations}

To accelerate the fixed point method, we can instead create a new
iterative scheme by approximating the operator $\vA$ using the Monte
Carlo method to estimate the solution error. With this idea hand, the
Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can be
defined as follows:

\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %% 
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:general-MCSA}
\end{subequations}

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

\subsection{Monte Carlo Method Selection}

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method outlined in
\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
adjoint method over the direct method, we choose the two-dimensional
time-dependent Poisson equation as a simple model problem:

\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}

@@

\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil both assuming a grid size of
$\Delta$ in both the $i$ and $j$ directions.

\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\
    + u_{i-1,j+1} + u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
  \label{eq:nine_point_stencil}
\end{multline}

For a single time step solution, we then have the following sparse
linear system to be solved with the MCSA method.

\begin{equation}
  \vA \vu^{n+1} = \vu^n
  \label{eq:poisson_eq_lin_sys}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ and was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permuations were simulated per
iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
solver. All computations were completed on a 2.66 GHz Intel Core i7
machine with 8 GB 1067 MHz DDR3 memory used for all computations
presented in this section. Fig.~\ref{fig:poisson_cpu_time} gives the
CPU time needed for each case to converge in seconds and
Fig.~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge relative to the problem size.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ mesh). Both the adjoint and direct solvers are used
    with the five point and nine point stencils. A significant CPU
    time speedup is noted with the adjoint method due to the reduced
    number of histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ mesh. Both the adjoint and direct solvers are used with
    the five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in several orders of magnitude
speedup over the direct solver while the number of iterations required
to converge is of a similar scale. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers and a system of size $N \times N$, the direct solver will
compute $N \times N$ random walks per permuation while the adjoint solver
will only compute one. Initially, this may not seem like a fair
comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time} that
the number of iterations requried to converge is approximately the
same. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, because the random walk
sequence contributes to the state in which it currently resides, fewer
histories are neccessary to compute contributions from all the nonzero
parts of the source. Therefore, although the total number of random
walks per iteration is in fact larger for the direct method, the
solution computed at each iteration is no better than that computed by
the adjoint method.

Although CPU time is shown to be greatly decreased by using the
adjoint solver in Fig.~\ref{fig:poisson_cpu_time}, the iteration
behavior was approximately the same. To assess the convergence
properties of MCSA using each solver and stencil, the infinity norm of
the residual computed in Eq.~(\ref{eq:general-MCSA}) was collected at
each iteration for a fixed problem
size. Fig.~\ref{fig:poisson_convergence} gives the results of these
computations. First, it is worthy to note on the semilog plot that we
are indeed achieving the expected exponential convergence from
MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior, leading to CPU time
being the dominating factor in method selection.

@@

We have presented a new Monte Carlo solution method for solving the
discrete, time-dependent radiation diffusion equation in 3D.  The MCSA
method has been shown to match results using standard solution
techniques to arbitrary precision.  Also, the new method has been to
shown to be faster than preconditioned CG and Richardson iteration for
the model problems presented here.
